Manual classification of tumor cells

### 2024-05-03 analyses/ewings-cell-type/01-marker-gene-tumor-classification.Rmd

- First we look at expression of each of the marker genes across all cells. 
- Then we use a z-transform prior to summing expression of all marker genes. 
Cells with a z-score for any marker gene > 0 are classified as tumor cells. 
- We look at marker genes for these tumor cells using `scran::findMarkers()` and compare to the tumor markers.
- Marker gene expression is examined across clusters and tumor and normal cells are classified using clusters rather than individual cells. 
- Finally, we compare the cells classified as tumor/normal manually to assignments from `SingleR` and `CellAssign`.

@@

```{r packages}
suppressMessages(
  # load required packages
  library(SingleCellExperiment),
  library(ggplot2)
)
```

@@

# The path to this module
module_base <- file.path(repository_base, "analyses", "ewings-cell-type")

# source in helper functions
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

@@

```{r}
# faceted umap showing a umap panel for each marker gene 
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.1, size = 0.2) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(
      color = "Log-normalized gene expression"
    ) +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
  theme_bw()
```

@@

```{r}
ggplot(umap_df, aes(x = gene_expression, fill = gene_symbol)) +
  geom_density() +
  theme_bw() +
  facet_wrap(vars(gene_symbol)) +
  theme(legend.position = "none")
```

@@

umap_df <- umap_df |> 
  dplyr::group_by(gene_symbol) |>
  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1] |>
  dplyr::ungroup()

@@

```{r}
ggplot(umap_df, aes(x = transformed_gene_expression, fill = gene_symbol)) +
  geom_density() +
  theme_bw() +
  facet_wrap(vars(gene_symbol)) +
  theme(legend.position = "none")
```

@@

# plot mean gene expression 
ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c() +
  theme_bw()
```
Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest marker gene expression. 
We would anticipate that these are most likely to be the tumor cells.

@@

ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

@@

# make plot with gene on y-axis, expression on x-axis and then color density by tumor/normal 
# do all marker genes so differences between tumor/normal or only a subset? 
ggplot(celltype_df, aes(x=transformed_gene_expression, color = sum_classification)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) +
  theme_bw()
```

@@

```{r}
# pull out top markers for tumor cells 
tumor_markers <- markers$Tumor |> 
  as.data.frame() |> 
  tibble::rownames_to_column("ensembl_gene_id") |> 
  dplyr::left_join(marker_genes_df) |>
  # just take those with FDR < 0.05 and logFC > 1.5
  dplyr::filter(FDR < 0.05, summary.logFC > 1.5) |> 
  dplyr::arrange(FDR)

@@

### Using clusters and marker gene expression to annotate tumor cells

We anticipate that the tumor cells and normal cells would cluster separately. 
First, let's just plot the cluster assignments alone.

These are clusters calculated in `scpca-nf` and its probable that these aren't the best clusters to use. 
Separately, we will evaluate clustering for these samples, but for now we will look at the clusters we have.

```{r}
# plot clusters on UMAP
ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = cluster)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

It looks like our large group of cells on the bottom right is comprised of a few clusters while most other groups of cells are a single cluster.

We can then look at marker gene expression across clusters.
Below we visualize the total number of cells in each cluster that are classified as tumor or normal using the sum of marker gene expression. 
Then for each set of cells, how many of those actually express each marker gene.

We expect to see two things:

- Tumor cells and normal cells tend to separate by cluster. 
- Good marker genes will be expressed in tumor cells but not normal cells.

```{r, fig.height=10, fig.width=10}
cluster_plot_df <- celltype_df |> 
  dplyr::mutate(gene_expresed = dplyr::if_else(transformed_gene_expression > 0, TRUE, FALSE)) |> 
  dplyr::select(gene_symbol, gene_expresed, sum_classification, cluster)

ggplot(cluster_plot_df, aes(x = sum_classification, fill = gene_expresed)) +
  geom_bar(stat = "count") +
  facet_grid(rows = vars(gene_symbol),
             cols = vars(cluster)) +
  scale_fill_manual(values = c("grey", "blue")) +
  theme_classic()

@@

Looking at this, most clusters are made up mostly of either normal or tumor. 
Also, genes like ADGRG2 and HES1 do not seem specific to tumor cells. There are normal and tumor cells that express the gene and not very many tumor cells actually have expression of ADGRG2. 
STEAP1 and NROB1 although not expressed in very many tumor cells, look to be specific to tumor cells which is promising.

CD99 and CCND1 are expressed in a high proportion of tumor cells but are also expressed in quite a few groups of normal cells. 
This is consistent with the literature for CD99 that it is expressed in normal cells, just at a lower level than in Ewing's cells.

Almost all genes have at least some expression in normal cells. 
Perhaps those cells should also be categorized as tumor cells, or they do have some baseline expression in normal cells which is possible.

Now we can classify the cells based on clustering and marker gene expression.

```{r}
# plot marker gene expression across all clusters 
ggplot(marker_sum_exp, aes(x = sum_exp, fill = cluster)) +
  geom_density() +
  facet_grid(rows = vars(cluster)) +
  theme_bw()
```

Looking at this there are definitely clusters with higher sum expression of the marker genes than other cells. 
If the cluster has a median expression > 0 then all cells in that cluster will get assigned as tumor cells.

```{r}
# classify clusters as tumor/normal 
cluster_classification_df <- marker_sum_exp |> 
  dplyr::group_by(cluster) |> 
  # calculate the median mean gene expression across all cells in a cluster
  dplyr::summarise(median_cluster_exp = median(sum_exp)) |> 
  # use 0.75 as the cutoff for tumor/normal 
  dplyr::mutate(cluster_classification = dplyr::if_else(median_cluster_exp > 0, "Tumor", "Normal"))

# get individual cell classifications
celltype_classification_df <- marker_sum_exp |> 
  dplyr::left_join(cluster_classification_df)

# umap showing tumor and normal cells 
ggplot(celltype_classification_df, aes(x = UMAP1, y = UMAP2, color = cluster_classification)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

@@

Looking at the heatmap in the supplemental report, the endothelial cells called by SingleR and CellAssign have a lot of overlap. 
Let's look at expression of marker genes in these endothlial cells compared to cells that are categorized as smooth muscle cells.

@@

ggplot(annotation_plot_df, aes(x = transformed_gene_expression, color = automated_annotation)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) +
  theme_bw()
```

@@

- We do see variation in marker gene expression across cells, suggesting that there are both tumor and normal cells here. - Using a z-score cut-off identifies a group of cells that appear to be tumor. 
- This mostly lines up with the default clustering that has been performed, where tumor and normal cells have been separated.
- The top marker genes for the "tumor" cells contain PRKCB and MAPT, two of the marker genes in our gene list. 
- It's likely that ADGRG2, HES1, NROB1, and STEAP1 will not be informative in assigning tumor cells. 
- We may be weary of genes like CD99 too as expression is more consistent across cells rather than specific to a group of cells. 
- Endothelial cells as identified by SingleR and CellAssign generally have low expression of these marker genes and muscle cells have higher expression of marker genes.

@@

```{r}
# get a dataframe with barcodes, sum_classification and cluster_classification 
celltype_classification_df <- celltype_classification_df |> 
  dplyr::select(
    barcodes,
    marker_gene_classification = sum_classification, 
    cluster_classification
  )

readr::write_tsv(celltype_classification_df, output_classifications_file)
```

>>>>

- First we look at expression of each of the marker genes across all cells. 
- Then we use a z-transform prior to summing expression of all marker genes. 
Cells with a z-score for any marker gene > 0 are classified as tumor cells. 
- We look at marker genes for these tumor cells using `scran::findMarkers()` and compare to the tumor markers.
- Marker gene expression is examined across tumor and normal cells. 
- Finally, we compare the cells classified as tumor/normal manually to assignments from `SingleR` and `CellAssign`.

@@

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(SingleCellExperiment)
  library(ggplot2)
})

# Set default ggplot theme
theme_set(
  theme_bw()
)
```

@@

# The path to this module
module_base <- file.path(repository_base, "analyses", "ewings-cell-type")
```

```{r}
# source in helper functions: make_jaccard_matrix and jaccard 
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

@@

```{r}
# faceted umap showing a umap panel for each marker gene 
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.1, size = 0.2) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(
      color = "Log-normalized gene expression"
    ) +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    theme(
      aspect.ratio = 1,
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) 
```

@@

We can also look at the distributions for each marker gene. 
I would expect to see some sort of bimodal distribution separating cells that do and do not have expression of the marker gene.

```{r}
ggplot(umap_df, aes(x = gene_expression, fill = gene_symbol)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) +
  theme(legend.position = "none")
```

Although we see some slight semblance of a bimodal distribution for most marker genes, it is hard to see and we cannot directly compare gene expression values across genes. 
This would make it hard to identify a cut off to categorize cells as expression or not expressing the marker gene.

@@

```{r}
umap_df <- umap_df |> 
  dplyr::group_by(gene_symbol) |>
  # get z-scores for each gene
  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |>
  dplyr::ungroup()
```

@@

```{r}
ggplot(umap_df, aes(x = transformed_gene_expression, fill = gene_symbol)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) +
  theme(legend.position = "none")
```

@@

# plot mean gene expression 
ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = sum_exp)) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_color_viridis_c()
```
Similar to the individual plots, it looks like there is one group of cells on the bottom right that has the highest marker gene expression. 
We would anticipate that these are most likely to be the tumor cells.

@@

ggplot(marker_sum_exp, aes(x = UMAP1, y = UMAP2, color = sum_classification)) +
  geom_point(size = 0.5, alpha = 0.5)
```

@@

# make plot with gene on y-axis, expression on x-axis and then color density by tumor/normal 
# do all marker genes so differences between tumor/normal or only a subset? 
ggplot(celltype_df, aes(x=transformed_gene_expression, color = sum_classification)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) 
```

@@

```{r}
# pull out top markers for tumor cells 
tumor_markers <- markers$Tumor |> 
  as.data.frame() |> 
  tibble::rownames_to_column("ensembl_gene_id") |> 
  # join to get gene symbols to print out later
  dplyr::left_join(marker_genes_df) |>
  # just take those with FDR < 0.05 and logFC > 1.5
  dplyr::filter(FDR < 0.05, summary.logFC > 1.5) |> 
  dplyr::arrange(FDR)

@@

### Looking at marker genes across tumor cells

We would expect that cells we are labeling as tumor cells have expression of the tumor marker genes. 
Cells labeled as normal should not have expression of the markers.

Below we visualize the total number of cells that are classified as tumor or normal using the sum of marker gene expression. 
Then for each set of cells, how many of those actually express each marker gene.

We expect that good marker genes will be expressed in tumor cells but not normal cells.

```{r, fig.height=10, fig.width=10}
marker_plot_df <- celltype_df |> 
  dplyr::mutate(gene_expresed = dplyr::if_else(transformed_gene_expression > 0, TRUE, FALSE)) |> 
  dplyr::select(gene_symbol, gene_expresed, sum_classification)

ggplot(marker_plot_df, aes(x = sum_classification, fill = gene_expresed)) +
  geom_bar(stat = "count") +
  facet_grid(rows = vars(gene_symbol)) +
  scale_fill_manual(values = c("grey", "blue")) +
  theme_classic()

@@

Looking at this might help give us an idea of genes that are more specifically expressed in tumor cells. 
Genes like ADGRG2 and HES1 do not seem specific to tumor cells. 
There are normal and tumor cells that express the gene and not very many tumor cells actually have expression of ADGRG2. 
STEAP1 and NROB1 although not expressed in very many tumor cells, look to be specific to tumor cells which is promising.

CD99, CAV1, and CCND1 are expressed in a high proportion of tumor cells but are also expressed in quite a few normal cells. 
This is consistent with the literature for CD99 that it is expressed in normal cells, just at a lower level than in Ewing's cells.

KDSR, MAPT, NKX2-2, PAPPA, and PRKCB all seem to have a high proportion of tumor cells expressing the gene and a low proportion of normal cells that express the gene. 
PRKCB and MAPT are also 2 of the top genes identified using the marker analysis.

Almost all genes have at least some expression in normal cells. 
Perhaps those cells should also be categorized as tumor cells, or they just have some baseline expression in normal cells which is possible.

@@

Looking at the heatmap in the supplemental report, the endothelial cells called by SingleR and CellAssign have a lot of overlap. 
Let's look at expression of marker genes in these endothelial cells compared to cells that are categorized as smooth muscle cells.

@@

ggplot(annotation_plot_df, aes(x = transformed_gene_expression, color = automated_annotation)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) 
```

@@

- We do see variation in marker gene expression across cells, suggesting that there are both tumor and normal cells here. - Using a z-score cut-off identifies a group of cells that appear to be tumor. .
- The top marker genes for the "tumor" cells contain PRKCB and MAPT, two of the marker genes in our gene list. 
- It's likely that ADGRG2, HES1, NROB1, and STEAP1 will not be informative in assigning tumor cells. 
- We may be weary of genes like CD99 too as expression is more consistent across cells rather than specific to a group of cells. 
- Endothelial cells as identified by SingleR and CellAssign generally have low expression of these marker genes and muscle cells have higher expression of marker genes.

@@

```{r}
# get a dataframe with barcodes, sum_classification and cluster_classification 
celltypes_df <- celltypes_df |> 
  dplyr::select(
    barcodes,
    marker_gene_classification = sum_classification
  )

readr::write_tsv(celltypes_df, output_classifications_file)
```
